---
title: "Benchmark with BayesPrism"
output: html_notebook
---

---
title: "Benchmark with BayesPrism"
output: html_notebook
---



```{r}
suppressWarnings(library(BayesPrism))
```

```{r}
library(anndata)
```




# TNBC 5 cell-type / 10 cell-type / Real data ST benchmark
### Load datasets
```{r}
# spatial dataset

# 5-cell-type simulation
#sim.path <- "data_spatial/simu_5_he/counts.st_synth.csv"
#st.exp <- read.csv("data_spatial/simu_5_he/counts.st_synth.csv", header = TRUE, row.names = 1)

# 10-cell-type simulation
st.exp <- read.csv("../data/simu_10/counts.st_synth.csv", header = TRUE, row.names = 1)

# real data
# st.path <- "../../starfysh/data/CID44971_TNBC/CID44971_TNBC.h5ad"
# adata.exp <- read_h5ad(st.path)
```


If benchmarking real data, convert `h5ad` object into dataframe with `spot` x `gene`:
```{r}
st.exp <- adata.exp$to_df()
colnames(st.exp) <- adata.exp$var[['_index']]
head(st.exp)
```


```{r}
# scRNA-seq dataset
sc.exp <- Matrix::readMM("../../starfysh/data/CID44971_TNBC/scrna/count_matrix_sparse.mtx")

# Load gene names & barcodes (without header)
gene.names = read.csv("../../starfysh/data/CID44971_TNBC/scrna/count_matrix_genes.tsv", sep = '\t', header = FALSE)$V1 
barcodes <- read.csv("../../starfysh/data/CID44971_TNBC/scrna/count_matrix_barcodes.tsv", sep = '\t', header = FALSE)$V1

sc.exp <- as.data.frame(t(as.matrix(sc.exp)))
rownames(sc.exp) <- barcodes
colnames(sc.exp) <- gene.names
head(sc.exp)
```

```{r}
# scRNA-seq annotation
celltype.map <- c(
    'Cancer Basal SC'=                  'Basal',
    'Cancer LumB SC'=                   'LumB',
    'Myoepithelial'=                    'Normal epithelial',
    'T_cells_c0_CD4+_CCR7'=             'Tcm',
    'T_cells_c1_CD4+_IL7R'=             'Tem',
    'T_cells_c3_CD4+_Tfh_CXCL13'=       'Tfh',
    'T_cells_c2_CD4+_T-regs_FOXP3'=     'Treg',
    'T_cells_c7_CD8+_IFNG'=             'Activated CD8',
    'T_cells_c8_CD8+_LAG3'=             'Terminal exhaustion',
    'B cells Memory'=                   'B cells memory',
    'B cells Naive'=                    'B cells naive',
    'Myeloid_c10_Macrophage_1_EGR1'=    'Macrophage M1',
    'Myeloid_c9_Macrophage_2_CXCL10'=   'Macrophage M2',
    'Plasmablasts'=                     'Plasmablasts',
    'Myeloid_c7_Monocyte_3_FCGR3A'=     'Monocytes',
    'Myeloid_c12_Monocyte_1_IL1B'=      'Monocytes',
    'Myeloid_c8_Monocyte_2_S100A9'=     'Monocytes',
    'Myeloid_c11_cDC2_CD1C'=            'cDC',
    'Myeloid_c3_cDC1_CLEC9A'=           'cDC',
    'Myeloid_c4_DCs_pDC_IRF7'=          'pDC',
    'CAFs MSC iCAF-like s1'=            'CAFs MSC iCAF-like',
    'CAFs MSC iCAF-like s2'=            'CAFs MSC iCAF-like',
    'CAFs myCAF like s4'=               'CAFs myCAF-like',
    'CAFs myCAF like s5'=               'CAFs myCAF-like',
    'PVL Immature s1'=                  'PVL immature',
    'PVL_Immature s2'=                  'PVL immature',
    'PVL Differentiated s3'=            'PVL differentiated',
    'Endothelial ACKR1'=                'Endothelial',
    'Endothelial CXCL12'=               'Endothelial',
    'Endothelial Lymphatic LYVE1'=      'Endothelial',
    'Endothelial RGS5'=                 'Endothelial'
)

annots.df <- read.csv("../../starfysh/data/CID44971_TNBC/scrna/metadata.csv", header = TRUE, row.names = 1)
head(annots.df)

```


```{r}
annots <- celltype.map[match(unlist(annots.df$celltype_subset), names(celltype.map))]
annots.df$celltype_subset <- annots
```


```{r}
table(annots.df$celltype_subset)
```



# Create cell type annotation mapping from Wu et al. to Starfysh



```{r}
is.sim = TRUE
if (is.sim == TRUE) {
  
  # 5 cell-type
  # mask <- (
  #   annots$celltype_major == 'CAFs' | 
  #   annots$celltype_major == 'Cancer Epithelial' | 
  #   annots$celltype_major == 'Myeloid' | 
  #   annots$celltype_major == 'Normal Epithelial' | 
  #   annots$celltype_major == 'T-cells'
  # )

  # subsetting annotations & original scRNA-seq matrix
  # annots.ss <- annots$celltype_major[mask]
  # sc.exp <- sc.exp[mask,]
  
  # 10 cell-type
  mask <- (
    (
      annots.df$celltype_subset == 'Activated CD8' | 
      annots.df$celltype_subset == 'B cells memory' | 
      annots.df$celltype_subset == 'Basal' | 
      annots.df$celltype_subset == 'CAFs MSC iCAF-like' | 
      annots.df$celltype_subset == 'CAFs myCAF-like' | 
      annots.df$celltype_subset == 'Endothelial' | 
      annots.df$celltype_subset == 'PVL immature' | 
      annots.df$celltype_subset == 'Tcm' | 
      annots.df$celltype_subset == 'Treg' | 
      annots.df$celltype_subset == 'pDC'  
    ) & (!is.na(annots.df$celltype_subset))
  )
  
  # subsetting annotations & original scRNA-seq matrix
  annots.ss <- annots.df$celltype_subset[mask]
  sc.exp.ss <- sc.exp[which(mask),]

} else {
  # For real data, eliminate unannotated entries
  mask <- !is.na(annots.df$celltype_subset)
  annots.ss <- annots.df$celltype_subset[mask]
  sc.exp.ss <- sc.exp[mask,]
}
```


```{r}
head(sc.exp.ss)
```

### BayesPrism QC matrices
Cell type / state annotations

```{r}
plot.cor.phi(
  input = sc.exp.ss,
  input.labels = annots.ss,
  min.exp = 1,
  title = "cell state correlation",
  cexRow = .2, cexCol = .2,
  margin = c(2, 2)
)
```

# Outlier gexp plots for ST & SC matrices
```{r}
sc.stat <- plot.scRNA.outlier(
  input = sc.exp.ss,
  cell.type.labels = annots.ss,
  species = "hs",
  return.raw = FALSE
)
```

```{r}
bk.stat <- plot.bulk.outlier(
  bulk.input = st.exp,
  sc.input = sc.exp.ss,
  cell.type.labels = annots.ss,
  species = "hs",
  return.raw = TRUE
)
```

Filter out outlier genes from scRNA-seq (including sex-chr & low-expr genes)
```{r}
sc.exp.filtered <- cleanup.genes(
  input = sc.exp.ss,
  input.type = "count.matrix",
  species = "hs",
  gene.group = c("Rb","Mrp","other_Rb","chrM","MALAT1","chrX","chrY"),
  exp.cells = 5
)

dim(sc.exp.filtered)
```

Visualize concordance of gexp for different gene types:
```{r}
plot.bulk.vs.sc(
  sc.input = sc.exp.filtered,
  bulk.input = st.exp,
)
```

Further subset with
- protein-coding genes from scRNA-seq (as shown above)
- DEGs for each cell type 
```{r}
sc.exp.filtered <- select.gene.type(sc.exp.filtered, gene.type = "protein_coding")

diff.exp.stat <- get.exp.stat(
  sc.dat = sc.exp.ss[,colSums(sc.exp>0)>3],
  cell.type.labels = annots.ss,
  cell.state.labels = annots.ss,
  psuedo.count = .1,
  cell.count.cutoff = 50,
  n.cores = 1
)

sc.exp.filtered <- select.marker(
  sc.dat = sc.exp.filtered,
  stat = diff.exp.stat,
  pval.max = .01,
  lfc.min = .1
)

dim(sc.exp.filtered)
```

### Deconvolution

Construct scRNA-seq `prism` object
```{r}
myPrism <- new.prism(
  reference = sc.exp.filtered,
  mixture = st.exp,
  input.type = "count.matrix",
  cell.type.labels = annots.ss,
  cell.state.labels = annots.ss,
  key = NULL, # treat all cell types equally, switch to see effects
  outlier.cut = .01,
  outlier.fraction = .1
)
```



```{r}
# Deconvolution
t0 <- Sys.time()
bp.res <- run.prism(prism = myPrism, n.cores = 4)
t1 <- Sys.time()
duration <- t1 - t0
```


```{r}
dim(theta)
```

```{r}
colnames(theta)
```


Extract deconvolution results
```{r}
theta <- get.fraction(
  bp = bp.res,
  which.theta = "final",
  state.or.type = "type"
)

write.csv(theta, "results/Bayesprism_proportion_simu_10.csv")
```


```{r}
duration
```
```{r}
cat("Runtime:", as.double(duration)*60, "minutes")
```








